Scalable Extreme Deconvolution

### 2019-09-01 ml4ps_paper/paper.tex
Update paper.

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage{neurips_2019}

@@

\newcommand{\bx}{\mathbf{x}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bm}{\mathbf{m}}
\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}

@@

The Extreme Deconvolution method is a mixture-model based approach originally intended to perform density estimation on noisy astronomy datasets with known noise covariances for each datapoint.
Newer much larger datasets are available to which the existing method for fitting the model cannot scale to.
We propose two extensions to the extreme deconvolution method based on an online variation of the EM algorithm, and direct optimisation of the log-likelihood via SGD, both which can make use of GPU computation.
We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.

@@

Many scientific applications require an estimate of some density $p(\bx)$ to be fitted using a set of samples $\bx_i$.
In many cases the dataset is noisy, but we wish to fit our density estimate to the true underlying noiseless data.
Extreme Deconvolution is a method that allows Gaussian Mixture Models (GMMs) to be fitted to data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
The method was originally developed to perform density estimation on the dataset produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
The Hipparcos catalogue consists of astrometric solutions (positions and velocities on the sky) and photometry (light intensity) for 118,218 stars, with associated noise covariances provided for each entry.

The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
The existing method of fitting extreme deconvolution models cannot cope with such a dataset, as it requires all of the data to fit in memory, and must make a full pass over the dataset before it can update parameters.
In this work, we propose two new methods for fitting the extreme deconvolution model that can scale to such datasets, based on an online variation of the Expectation-Maximisation (EM) algorithm, and stochastic gradient descent (SGD) on the log-likelihood.
We develop implementations that can leverage GPU-based computation, and show that they provide comparable density estimates to the existing method, whilst being much faster to train.

@@

The aim of extreme deconvolution is to perform density estimation on a noisy $d$-dimensional dataset $\{\bx_i\}_{i=0}^N$, where $\bx_i$ was generated by adding zero-mean Gaussian noise $\epsilon_i$ with known per-datapoint covariance $S_i$ to a true value $\bv_i$,
\begin{equation}
  \mathbf{x_i} = \bv_i + \epsilon_i,\quad  \epsilon_i \sim \mathcal{N}(\mathbf{0}, S_i).
\end{equation}
We assume that $\bv_i$ can be modelled by a mixture of Gaussians with $K$ components,
\begin{equation}
p(\bv_i \mid \theta) = \sum_j^K \alpha_j \mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K
\end{equation}
As both the noise model is Gaussian and the density estimation model is a mixture of Gaussians, the likelihood of $\bx_i$ is also a Gaussian mixture,
and the total log-likehood of the dataset is
\begin{equation}
\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\mathcal{N}(\bv_i \mid \mathbf{m}_j, T_{ij}), \quad T_{ij} = V_j + S_i.
\end{equation}
The original model also includes a method for handling the case where $\bx_i$ has been transformed from $\bv_i$ via a per-datapoint projection matrix $R_i$, which we omit here for simplicity.
Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.

@@

\subsection{Online Expectation-Maximisation}
The original method for fitting extreme deconvolution models used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
In the E-step, the expected sufficient statistics for each datapoint that need to be computed are $s_{ij} = \{q_{ij}$, $q_{ij}\bv_i$, and $q_{ij}\bv_i\bv_i^T\}$.
The M-step comprises normalisation of the sums of expected sufficient statistics to get the mixture component parameters.

\citet{cappeOnlineExpectationMaximization2009} proposed an online variation of EM for latent data models.
Their method involves computing a stochastic approximation to the sums of expected sufficient statistics.
At each iteration $t$, the sums of expected sufficient statistics $s_t = \sum_i s_{ij}$ are computed over a minibatch using the current estimate of the parameters, and the stochastic estimate $\hat{s}_t$ updated as
\begin{equation}
 \hat{s}_{t} = (1 - \lambda)\hat{s}_{t-1} + \lambda s_t
\end{equation}
where $\lambda$ is a sufficiently small step-size.
$\lambda$ should theoretically decrease at each iteration according to the Robbins-Monro conditions, but in practice we found that using a constant step-size was sufficient.
After each update, the parameters of the model are updated by normalising $\hat{s}_t$ appropriately.

\todo{More complete explanation if space.}

A direct implementation suggests that we should update $V_j$ as
\begin{equation}
V_j  = \mathrm{E}(q_{ij} \bv_ij \bv_ij^T) - \bm_j \bm_j^T
\end{equation}
This is usually acceptable when using double-precision floats, but using GPU based computation restricts us to single precision.
Performing the update of $V_j$ like this with single-precision floats will result in catastrophic cancellation if the means are sufficiently large compared to the variance.
Instead, we compute smoothed versions of the sum of squared deviation products which avoids this issue.
\begin{equation}
\end{equation}

\todo{Better justification of why this works, or alternate method.}

@@

An alternative to EM-based methods is to optimise the log-likelihood directly.
The log-likelihood has some constraints on its parameters, namely that the weight $a_j$ sum to $1$ and that the covariances $V_j$ are positive definite.
Directly fitting it with standard gradient-based optimisers requires a transformation of the parameters to remove the constraints.
The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$,
\begin{equation}
\alpha_j = \frac{e^{z_j}}{\sum_k^K e^{z_k}}.
\end{equation}
We parametrise the covariances $V_j$ via the Cholesky decomposition, 
\begin{equation}
  V_j = L_jL_j^T
\end{equation}
where $L_j$ is a lower-triangular matrix.
We also constrain the diagonal elements of $L_j$ to be positive by taking the log of them.
Note that for a standard GMM we could bypass forming the covariance altogether and evaluate the log PDF of the Normal distribution directly with the Cholesky factor.
For the extreme deconvolution model, this cannot be done, as we need to form the covariance $Tij$ for every datapoint.
Having removed the constraints, we can then take the gradient of the log-likelihood with respect to the parameters and optimise it using a standard minibatch-based optimiser.
Pre-multiplying the gradient with the inverse Fisher information matrix may be necessary to keep the covariances stable during optimisation with standard stochastic gradient descent.
In practice we found that using the Adam optimisation algorithm eliminates the need to do this~\cite{kingmaAdamMethodStochastic2014}.

@@

We implemented both of our methods in PyTorch.
The use of PyTorch allows us to take advantage of its automatic differentiation framework and in-built optimisers for the SGD method,
as well as allowing us to write code targeting both CPU and GPU-based computation.

To evaluate our methods, we used a subset of rows from the Gaia DR2 source table \cite{brownGaiaDataRelease2018}.
We selected the 5 primary astrometric features, along with the BP-RP colour and mean flux in the G-band.
In total there were approximately 2 million rows.
Where data were missing, we set the field to zero and the noise variance to a large value.
This is only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
We used the same number of components $K=128$ for each model, following \cite{andersonImprovingGaiaParallax2018}.
In practice we would want to select a value of $K$ by cross-validation.

Initialisation of all models was performed by first running a minibatch version of the K-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
Initial covariances were set to the identity matrix.
Early stopping was done based on a held-out validation set comprising 10\% of the rows.
Final model performance was evaluated on another held-out test set comprising another 10\% of the rows.
Table~\ref{results-table} reports the validation and test log-likelihoods for each method.

\begin{table}
  \caption{Total validation and test log-likelihoods along with average training time for the Gaia data subset. Average over 5 runs with standard deviation.}
  \label{results-table}
  \centering
  \begin{tabular}{lccc}
    \toprule
    Method     & Validation     & Test & Training Time (s) \\
    \midrule
    Existing EM & -  & -     \\
    Online EM     & - & -      \\
    SGD     & $-5.468 \times 10^6 \pm 2.46 \times 10^3$        & - & $1469 \pm 11$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Further improvements to our approaches are possible.
The original paper presents a method of getting out of local maxima by splitting and merging clusters, with the split-merge criteria evaluated on the whole dataset.
It should be possible to replace the criteria with stochastic estimates, which would permit them to be used with both the SGD and online EM methods.
Our approaches also add more free parameters to be selected, including learning rate and batch size.
This adds scope for hyperparameter optimisation to improve the log-likelihood, potentially making use of Bayesian optimisation.

The question remains of which of our two methods should be used for any particular problem?
As the results show, the SGD method is faster than the online EM method, at a cost of having a slightly worse log-likelihood.
The online EM method performed better, but during training would occasionally terminate early as a result of an update forming an invalid covariance matrix, necessitating a restart.
(This issue also exists with the implementation of the existing method).
The answer will therefore depend on exactly what dataset we wish to perform density estimation on, and how accurate an estimation we require.
In practice, we would suggest using the online EM if the size of the dataset was small enough to permit several restarts, whilst an extremely large dataset would probably be better fitted with the SGD method.

\todo{Add an example plot of online-EM going wrong if space?}

>>>>

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[preprint]{neurips_2019}

@@

\newcommand{\bx}{\mathbf{x}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bm}{\mathbf{m}}
\newcommand{\bb}{\mathbf{b}}

\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}

@@

The Extreme Deconvolution method fits a probability density to a dataset where every observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomy datasets.
The existing method of fitting the model cannot scale to new larger astronomy datasets.
We propose two extensions to the extreme deconvolution method based on an online variation of the EM algorithm, and direct optimisation of the log-likelihood via minibatch gradient descent, both of which can make use of GPU computation.
We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.

@@

Extreme deconvolution is a method that fits Gaussian Mixture Models (GMM)s to noisy data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
The method was originally developed to perform probabilistic density estimation on the dataset of stellar velocities produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
The Hipparcos catalogue consists of astrometric solutions (positions and velocities on the sky) and photometry (light intensity) for 118,218 stars, with associated noise covariances provided for each entry.

The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
Previous work using an extreme deconvolution model on the Gaia catalogue had to work with a subset of the data, as the existing method of fitting the models cannot work with very large datasets \cite{andersonImprovingGaiaParallax2018}.
It requires all of the data to fit in memory, and must make a full pass over the dataset before it can update parameters.

In this work, we propose two new methods for fitting the extreme deconvolution model that can scale to the full dataset size, based on an online variation of the Expectation-Maximisation (EM) algorithm, and minibatch gradient descent on the log-likelihood.
We develop implementations that can leverage GPU-based computation, and show that they provide comparable density estimates to the existing method, whilst being much faster to train.

@@

The aim of extreme deconvolution is to perform density estimation on a noisy $d$-dimensional dataset $\{\bx_i\}_{i=0}^N$, where $\bx_i$ was generated by adding zero-mean Gaussian noise $\epsilon_i$ with known per-datapoint covariance $S_i$ to a projection $R_i$ of a true value $\bv_i$,
\begin{equation}
  \bx_i = R_i\bv_i + \epsilon_i,\quad  \epsilon_i \sim \mathcal{N}(\mathbf{0}, S_i).
\end{equation}
We assume that $\bv_i$ can be modelled by a mixture of Gaussians with $K$ components,
\begin{equation}
p(\bv_i \mid \theta) = \sum_j^K \alpha_j \mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K
\end{equation}
As both the noise model is Gaussian and the density estimation model is a mixture of Gaussians, the likelihood of $\bx_i$ is also a Gaussian mixture,
and the total log-likehood of the dataset is
\begin{equation}
\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij}), \quad T_{ij} = R_iV_jR_i^T + S_i.
\end{equation}

Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.
For simplicity in the remainder of this work we will set $R_i$ to be the identity matrix and handle missing data via the covariance method.

@@

\subsection{Minibatch Expectation-Maximisation}

The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size N, using our current estimate of the parameters,
\begin{equation}
r_{ij} = \frac{\alpha_j \mathcal{N}(\bx_i \mid \bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid \bm_k, T_{ik})}, \quad
\bb_{ij} = m_j + V_jT_{ij}^{-1}(\bx_i - \bm_j), \quad
B_{ij} = V_j - V_j T_{ij}^{-1}V_j.
\end{equation}
The $r_{ij}$ term is just the posterior probability of datapoint $\bx_i$ coming from component $j$.
The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
The expected sufficient statistics are then summed together over the minibatch,
\begin{equation}
\phi_j^1 = \sum_i r_{ij}, \quad
\phi_j^2 = \sum_i r_{ij} \bb_{ij}, \quad
\phi_j^3 = \sum_i r_{ij} [\bb_{ij}\bb_{ij}^T + B_{ij}].
\end{equation}
Stochastic estimates $\hat{\phi}_{tj} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
\begin{equation}
\hat{\phi}_{tj} = (1 - \lambda)\hat{\phi}_{(t-1)j} + \lambda \phi_{tj}.
\end{equation}
Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters.
\begin{equation}
\alpha_j = \frac{\hat{\phi}_j^1}{N}, \quad
\bm_j = \frac{\hat{\phi}_j^2}{\hat{\phi}_j^1}, \quad
V_j = \frac{\hat{\phi}_j^3}{\hat{\phi}_j^1} - \bm_j \bm_j^T.
\end{equation}
This procedure is repeated with new randomly ordered minibatches until convergence.
If we set $\lambda=1$ and replace each minibatch with the entire dataset then we recover the original fitting method.

\todo{Discuss numerical stability of the $V_j$ update.}

@@

An alternative to EM-based methods is to optimise the log-likelihood directly.
The log-likelihood has some constraints on its parameters, namely that the weight $a_j$ sum to $1$ and that the covariances $V_j$ are positive definite.
Directly fitting it with standard gradient-based optimisers requires a transformation of the parameters to remove the constraints.
The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$, and the covariances $V_j$ as a lower triangular matrix $L_j$ via the Cholesky decomposition, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of the unconstrained element $\tilde{L}_q$,
\begin{equation}
\alpha_j = \frac{e^{z_j}}{\sum_{k=1}^K e^{z_k}}, \quad
V_j = L_jL_j^T, \quad
(L_j)_{qq} = \exp({\tilde{L}_q}).
\end{equation}
Note that for a standard GMM we could bypass forming the covariance altogether and evaluate the log PDF of the Normal distribution directly with the Cholesky factor.
For the extreme deconvolution model, this cannot be done, as we need to form the covariance $Tij$ for every datapoint.
Having removed the constraints, we can then take the gradient of the log-likelihood with respect to the parameters and optimise it using a standard minibatch-based optimiser.

@@

We implemented both of our methods in PyTorch.
To evaluate our methods, we used a subset of rows from the Gaia DR2 source table \cite{brownGaiaDataRelease2018}.
We selected the 5 primary astrometric features, along with the BP-RP colour and mean flux in the G-band.
In total there were approximately 2 million rows.
Where data were missing, we set the field to zero and the noise variance to a large value.
This is only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
We used the same number of components $K=128$ for each model, following \cite{andersonImprovingGaiaParallax2018}.
In practice we would want to select a value of $K$ by cross-validation.

Initialisation of all models was performed by first running a minibatch version of the K-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
Initial covariances were set to the identity matrix.
Early stopping was done based on a held-out validation set comprising 10\% of the rows.
Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection performed based on this value.
Table~\ref{results-table} reports the validation and test log-likelihoods for each method.

\todo{Add a 2D projection of an example solution?}

\begin{table}
  \caption{Total validation and test log-likelihoods along with average training time for the Gaia data subset. Average over 5 runs with standard deviation.}
  \label{results-table}
  \centering
  \begin{tabular}{lccc}
    \toprule
    Method     & Validation     & Test & Training Time (s) \\
    \midrule
    Existing EM & -  & -     \\
    Minibatch EM     & - & -      \\
    SGD     & $-5.468 (\pm 0.0025) \times 10^6$        & - & $1469 \pm 11$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Further improvements to our approaches are possible.
The original paper presents a method of getting out of local maxima by splitting and merging clusters, with the split-merge criteria evaluated on the whole dataset.
It should be possible to replace the criteria with stochastic estimates, which would permit them to be used with both the SGD and minibatch EM methods.
Our approaches also add more free parameters to be selected, including learning rate and batch size.
This adds scope for hyperparameter optimisation to improve the log-likelihood, potentially making use of Bayesian optimisation.

The question remains of which of our two methods should be used for any particular problem?
As the results show, the SGD method is faster than the minibatch EM method, at a cost of having a slightly worse log-likelihood.
The minibatch EM method performed better, but during training would occasionally terminate early as a result of an update forming an invalid covariance matrix, necessitating a restart.
(This issue also exists with the implementation of the existing method).
The answer will therefore depend on exactly what dataset we wish to perform density estimation on, and how accurate an estimation we require.
In practice, we would suggest using the minibatch EM if the size of the dataset was small enough to permit several restarts, whilst an extremely large dataset would probably be better fitted with the SGD method.

\todo{Add discussion of complexity}
\todo{Add discussion of projections $R_{ij}$}
